Interpreting the gas-fraction as the entropy of an isothermal dusty gas

### 2016-07-29 model.tex
polishing up model section

\subsection{Effective energy equation}
We show that for a prescribed temperature distribution, the mixture
obeys an evolutionary energy equation, due to the advection of the
dust-fraction. Eliminate $\epsilon$ in favor of $P$, 
\begin{align*}
  \epsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional source 
terms due to the imposed temperature profile, as well as dust-gas
friction.

Eq. \ref{eff_energy} can be written in conservative form,
\begin{align*}
  \frac{\p P}{ \p t} + \nabla\cdot\left\{P\left[\bm{v} -
      \tstop\left(c_s^2 - \frac{P}{\rho}\right)\nabla\ln{P}\right]
    \right\}\\
  = \left[P\bm{v} - \tstop\left(c_s^2 - \frac{P}{\rho}\right)\nabla
    P\right]\cdot\nabla\ln{c_s^2}. 
\end{align*}
We may thus re-interpret $P$ as the mixture's energy density, but the
energy flux has an additional contribution from the pressure
gradient. The term on the right-hand-side, owing to the imposed
temperature profile, can be interpreted as an external heat source.

\section{Equilibrium}\label{eqm}
To obtain an axisymmetric steady state with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, 
we need to solve 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}\\
  0 & = \nabla\cdot\left(\epsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
for $\rho$, $\Omega$ and $\epsilon$ with $P=P(\epsilon,\rho)$ given by
the equation of state (Eq. \ref{eos}).

Eq. \ref{steady_dust} makes it difficult to obtain equilibrium
solutions explicitly. However, if the diffusive process is slow and
can be neglected, then one may just solve
Eq. \ref{steady_momr}---\ref{steady_momz} with a prescribed (initial)
distribution of the dust fraction $\epsilon(r,z)$.

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
We assume a Gaussian profile in the dust-to-gas ratio, 
\begin{align}\label{dust_gauss}
  \tepsilon(r,z) = \tepsilon_0(r)
  \exp{\left[-\frac{z^2}{2\widetilde{H}^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\tepsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \tepsilon_0(r) = \tepsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\tepsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  \rhog(r,z) = \rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\tepsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

@@

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profil (Eq. \ref{power_temp}) e gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{
  \frac{\p\tepsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\tepsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{\right\} }
  \left. +\frac{q}{r}\left[\frac{\p\tepsilon}{\p z} + \left(1+\tepsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

@@

\subsection{Strictly isothermal gas perfectly coupled to dust}  
When $c_s^2$ is a constant and $\tstop=0$, the dusty gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit adiabatic
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\epsilon$ is advected with
the gas (i.e. evolves like an entropy when there is no heating or
cooling).

In this case one may immediately apply the Solberg-Hoiland
criteria to assess axisymmtric \emph{stability}:

\begin{align}
  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0,\\
  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, 
\end{align}
with $s = \ln{P/\rho^\gamma}$ and 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the
square of the epicylic frequency. We assume $\kappa^2>0$. 
These criteria are applicable to any adiabatic fluid. For 
isothermal gas perfectly coupled to dust, we may take 
$s=\ln{(1-\epsilon)}$. 
%from the definition of $s$ and  
%the equation of state (Eq. \ref{eos}) with uniform sound-speed. 
%\begin{align}
%  s \equiv \ln \frac{P}{\rho} = \ln{(1-\epsilon)} 
%\end{align}
% The first equality is
%similar to the usual defintion for gas entropy (with unit adiabatic
%index). The second follows from the imposed equation of state,
%Eq. \ref{eos}, with a constant sound-speed. 
The stability conditions are thus
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\epsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\epsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \epsilon}{\p r} \right) & > 0. \label{dusty_solberg2}
\end{align}

%In accretion disks, $\rhog$, $\rhod$ and $\epsilon$ are expected to
%decrease away from the mid-plane
%\begin{align*}  
%  \frac{\p\rhog}{\p |z|} < 0,
%, \quad \frac{\p \rhog}{\p r} < 0,   
%\end{align*}
%and similarly for $\rhod$, $\epsilon$, and the dust-to-gas ratio $\tepsilon$. 
%We may thus evaluate the stability criteria at $z>0$ without loss of
%generality.

@@

\subsubsection{Order-of-magnitude estimates} 
We evaluate the above stability criteria for the vertically-Gaussian
distributions for the gas density and the dust-to-gas ratio as 
described in \S\ref{eqm}. However, here we permit 
$\tepsilon_0=\tepsilon_0(r)$ and $\Htilde=\Htilde(r)$ to vary with
radius. We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$. In terms of the dust-to-gas
ratio $\tepsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\tepsilon}{\left(1+
    \tepsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} + 
  \frac{z^2}{\widetilde{H}^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\tepsilon
  h_\mathrm{g}^2}{\left(1+\tepsilon\right)^2}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\widetilde{H}^2}{H_\mathrm{g}^2}\frac{\p\ln{\tepsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$ and
$h_\mathrm{g},\tepsilon, \widetilde{H}/H_\mathrm{g}\ll 1$, since then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.

@@

\subsection{Locally isothermal gas perfectly coupled to dust} 
When $c_s(r,z)$ is non-uniform but $\tstop=0$, simple stability
criteria are not available. However, we may obtain an integral
relation for linear, axisymmetric waves with frequency $\sigma$ (see
Appendix \ref{var_prin}, where we also account for finite $\tstop$) in
the form:   
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV - \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.
\end{align}
The real coefficients $A,B,C,D$ can be read off
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}).

If $c_s$ is constant, then $\sigma^2$ is real, and 
the first integral leads to the previous Solberg-Hoiland criteria for 
axisymmetric stability. However, when $c_s$ is non-uniform, $\sigma^2$ 
is generally complex, implying instability. This is responsible for 
the usual vertical shear instability in locally isothermal disks.

\subsection{Strictly isothermal gas well-coupled to dust} 
Now consider $c_s^2=\mathrm{const.}$ but $\tstop\neq0$. In this case 
the effective energy equation is 
\begin{align}\label{dusty_cooling}
  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
\end{align} 
The function $\mathcal{C}$ can be considered as a cooling term. By
analogy with enabling VSI in an adiabatic gas through cooling
\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
if this term is large enough (physically meaning diffusion of the
dust-fraction is strong).

Note that $\mathcal{C}$ is also responsible for the streaming
instability  in the strong-drag limit \citep{laibe14}, though this
does not require vertical stratification. In the form of
Eq. \ref{dusty_cooling}, however, the streaming instability has a
thermodynamical interpretation, similar to thermal instabilities in
stars \citep{latter06}.

>>>>

\subsection{Effective energy equation}
We show that for a prescribed temperature distribution, the mixture
obeys an evolutionary energy equation, due to the advection of the
dust-fraction. Eliminate $\epsilon$ in favor of $P$, 
\begin{align*}
  \epsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional energy source 
terms due to the imposed temperature profile via $\nabla c_s^2$, as
well as dust-gas drag via $\mathcal{C}$. 
%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.

Eq. \ref{eff_energy} can be written in conservative form,
\begin{align*}
  \frac{\p P}{ \p t} + \nabla\cdot\left\{P\left[\bm{v} -
      \tstop\left(c_s^2 - \frac{P}{\rho}\right)\nabla\ln{P}\right]
    \right\}\\
  = \left[P\bm{v} - \tstop\left(c_s^2 - \frac{P}{\rho}\right)\nabla
    P\right]\cdot\nabla\ln{c_s^2}. 
\end{align*}
We may thus re-interpret $P$ as the mixture's energy density, but the
energy flux has an additional contribution from the pressure
gradient and dust-gas friction. 
The term on the right-hand-side, owing to the imposed
temperature profile, can be interpreted as an external heat source.

\section{Dynamical equilibria}\label{eqm}

For a given distribution of the dust-fraction $\epsilon$ (or
dust-to-gas ratio $\tepsilon$), the 
mass and momentum Eqs. \ref{masseq}---\ref{momeq} admit     
axisymmetric steady states with $\rho(r,z)$ and 
$\bm{v}=r\Omega(r,z)\hat{\bm{\phi}}$ where $\Omega = v_\phi/r$, which satisfy 
\begin{align}
  r\Omega^2 &= \frac{\p \Phi}{\p r} + \frac{1}{\rho}\frac{\p P}{\p
    r},\label{steady_momr}\\
  0 & = \frac{\p\Phi}{\p z} + \frac{1}{\rho}\frac{\p P}{\p z},\label{steady_momz}
%  0 & = \nabla\cdot\left(\epsilon\tstop\nabla P\right) \label{steady_dust}
\end{align}
with $P=P(\epsilon,\rho)$ given by the equation of state
(Eq. \ref{eos}). An explicit solution is presented in
\S\ref{steady_state}.  
However, disk structures obtained this way generally do not satisfy 
the steady state effective energy Eq. \ref{eff_energy} 
except in the limit of perfectly coupled dust ($\tstop\to0$).

True steady states may be calculated by invoking an underlying gas
turbulence, which leads to dust diffusion \citep{takeuchi02, youdin07, 
  lyra13}. This complication is beyond the scope of this paper and we 
defer it to a future work. Here, we will either consider $\tstop = 0$,
or assume that $\tstop$ is sufficiently small such that the background
disk (as obtained from Eq. \ref{steady_momr}---\ref{steady_momz}) does 
not evolve significantly.  %relative to perturbations

%Eq. \ref{steady_dust} makes it difficult to obtain equilibrium
%solutions explicitly. However, if the diffusive process is slow and
%can be neglected, then one may just solve
%Eq. \ref{steady_momr}---\ref{steady_momz} with a prescribed (initial)
%distribution of the dust fraction $\epsilon(r,z)$.

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
To obtain an actual disk structure for numerical computations, we
assume a Gaussian profile in the dust-to-gas ratio,   
\begin{align}\label{dust_gauss}
  \tepsilon(r,z) = \tepsilon_0(r)
  \exp{\left[-\frac{z^2}{2\widetilde{H}^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\tepsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \tepsilon_0(r) = \tepsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\tepsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  &\rhog(r,z)= \notag\\
&\rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\tepsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

@@

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profile (Eq. \ref{power_temp}) gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{
  \frac{\p\tepsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\tepsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{\right\} }
  \left. +\frac{q}{r}\left[\frac{\p\tepsilon}{\p z} + \left(1+\tepsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

@@

In this section we review the basic properties of the 
single-fluid mixture using Eqs.\ref{masseq}--\ref{momeq} with
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding linearized equations about an
axisymmetric steady state. These lead to the integral relation,

\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV  -\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV \notag\\
&\phantom{===}
- \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV,\label{int_rel}
\end{align}
between the (complex) mode frequency $\sigma$ and the spatial
structure of the disturbances which are prefixed by deltas. 
The real coefficients $A,B,C,D$ can be read off
Eq. \ref{integral_ex}. Note that $B=C$ from dynamical equilibrium
(Eq. \ref{vshear}).

\subsection{Strictly isothermal gas perfectly coupled to dust}   
When $c_s^2$ is a constant and $\tstop=0$, the dusty-gas equations are
exactly equivalent to that for adiabatic hydrodyamics with unit adiabatic
index. Although the gas is strictly isothermal, the mixture behaves 
adiabatically because the dust fraction $\epsilon$ is advected with 
the gas. This is similar to entropy being conserved following an
adiabatic gas when there is no heating or cooling.

In this case, the last two integrals in Eq. \ref{int_rel} vanish. Then 
requiring $\sigma^2>0$  gives the Solberg-Hoiland criteria 
for the axisymmtric \emph{stability} of an isothermal gas perfectly 
coupled to dust,  
\begin{align}
  \kappa^2 + c_s^2 \nabla\ln{\rhog}\cdot\nabla\epsilon &> 0,\label{dusty_solberg1}  \\
  -c_s^2\frac{\p\ln{\rhog}}{\p z}\left(-\kappa^2\frac{\p\epsilon}{\p
    z} + r\frac{\p\Omega^2}{\p z}\frac{\p \epsilon}{\p r} \right) & > 0, \label{dusty_solberg2}
\end{align}
%\begin{align}
%  \kappa^2 - \frac{1}{\rho}\nabla P \cdot \nabla s &> 0,\\
%  -\frac{1}{\rho}\frac{\p P}{\p z} \left(\kappa^2 \frac{\p s}{\p z} -
%  r\frac{\p\Omega^2}{\p z}\frac{\p s}{\p r}\right) &>0, 
%\end{align} 
where %$s = \ln{P/\rho}$, 
$\kappa^2 \equiv r^{-3}\p_r\left(r^4\Omega^2\right)$ is the
square of the epicylic frequency, and we assume $\kappa^2>0$. 
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} are equivalent to the
Solberg-Hoiland criteria for adiabatic gas if one writes the 
entropy $s=\ln{\left[c_s^2\left(1 - \epsilon\right)\right]}$.

%These criteria are applicable to any adiabatic fluid. 
%For an isothermal mixture, we may take $s=\ln{(1-\epsilon)}$.   
%from the definition of $s$ and  
%the equation of state (Eq. \ref{eos}) with uniform sound-speed. 
%\begin{align}
%  s \equiv \ln \frac{P}{\rho} = \ln{(1-\epsilon)} 
%\end{align}
% The first equality is
%similar to the usual defintion for gas entropy (with unit adiabatic
%index). The second follows from the imposed equation of state,
%Eq. \ref{eos}, with a constant sound-speed. 
%The stability conditions are thus
%In accretion disks, $\rhog$, $\rhod$ and $\epsilon$ are expected to
%decrease away from the mid-plane
%\begin{align*}  
%  \frac{\p\rhog}{\p |z|} < 0,
%, \quad \frac{\p \rhog}{\p r} < 0,   
%\end{align*}
%and similarly for $\rhod$, $\epsilon$, and the dust-to-gas ratio $\tepsilon$. 
%We may thus evaluate the stability criteria at $z>0$ without loss of
%generality.

@@

\subsubsection{Order-of-magnitude estimates} 
We evaluate the above stability criteria for the vertically-Gaussian
distributions for the gas density and the dust-to-gas ratio as 
described in \S\ref{eqm}. 
%However, here we permit 
%$\tepsilon_0=\tepsilon_0(r)$ and $\Htilde=\Htilde(r)$ to vary with
%radius. 
We assume the disk is approximately Keplerian so that
$\kappa\simeq\OmK$. In terms of the dust-to-gas
ratio $\tepsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\tepsilon}{\left(1+
    \tepsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}} + 
  \frac{z^2}{\widetilde{H}^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\tepsilon
  h_\mathrm{g}^2}{\left(1+\tepsilon\right)^2}
  \frac{\p\ln{\tepsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\widetilde{H}^2}{H_\mathrm{g}^2}\frac{\p\ln{\tepsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

Eq. \ref{stability_est1}---\ref{stability_est2} are easily satisfied
in thin accretion disks where radial gradients are $O(1/r)$, 
$h_\mathrm{g},\tepsilon,\ll 1$ and $
\widetilde{H}/H_\mathrm{g}\lesssim 1$. Then the
magnitude of the second term on the left-hand-side  of either
inequality is much less than unity. Thus radially smooth, strictly
isothermal disks perfectly coupled with dust are stable against
axisymmetric perturbations.

@@

\subsection{Locally isothermal gas perfectly coupled to dust} 
When $c_s(r,z)$ is non-uniform but $\tstop=0$, Eq. \ref{int_rel}
indicates that $\sigma^2$ is generally complex, implying instability.  
Writing $\sigma = 
\omega + \ii s$ where $\omega$ and $s$ are the real frequency and
growth rates, respectively, we have 
\begin{align}
  s = -\frac{\imag\int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}. 
\end{align} 
This is classic VSI caused by vertical shear arising from a radial
temperature gradient \citep{nelson13,barker15,lin15}.

%This is responsible for 
%the usual vertical shear instability in locally isothermal disks.
%However, we may obtain an integral
%relation for linear, axisymmetric waves with frequency $\sigma$ (see
%Appendix \ref{var_prin}, where we also account for finite $\tstop$) in
%the form:   
%If $c_s$ is constant, then $\sigma^2$ is real, and 
%the first integral leads to the previous Solberg-Hoiland criteria for 
%axisymmetric stability. However, when $c_s$ is non-uniform,

\subsection{Strictly isothermal gas well-coupled to dust} 
Now consider constant $c_s$ but $\tstop\neq0$. Then Eq. \ref{int_rel}
gives 
\begin{align}
  s = -\frac{\imag\int \left(\nabla\cdot\dd\bm{v}^*\right)\dd\mathcal{C}dV}{2\omega\int\rho\left(|\dd
    v_r|^2 + |\dd v_z|^2\right)dV}, 
\end{align}
%In this case 
%the effective energy equation is 
%\begin{align}\label{dusty_cooling}
%  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
%\end{align} 
%The function $\mathcal{C}$ can be considered as a cooling term. 
%By
%analogy with enabling VSI in an adiabatic gas through cooling
%\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
%if this term is large enough (physically meaning diffusion of the
%dust-fraction is strong). 
%Note that $\mathcal{C}$ is also responsible for the streaming
%instability  in the strong-drag limit \citep{laibe14}, though this
%does not require vertical stratification. In the form of
%Eq. \ref{dusty_cooling}, however, the streaming instability has a
%thermodynamical interpretation, similar to thermal instabilities in
%stars \citep{latter06}.  
which represent all instabilities due to large-but-finite dust-gas
friction. The formalism we have adopted give these instabilities a 
thermodynamical interpretation \citep[thermal instabilities in
  stars][]{latter06} because the dust-gas friction enters the problem
as an energy source term. This includes the streaming instability in
the strong-drag limit \citep{youdin05a, jacquet11, laibe14}.
